HTW EXAM Fits and Transfer Predictions

### 2023-11-12 _working/_htw_exam.qmd
model fitting

@@

| Step                          | Equation                                                                            | Description                                                                                                                            |
|------------------|------------------------|-------------------------------|
| **ALM Activation & Response** |                                                                                     |                                                                                                                                        |
| Input Activation              | $a_i(X) = \frac{e^{-c(X-X_i)^2}}{\sum_{k=1}^M e^{-c(X-X_k)^2}}$                     | Activation of each input node $X_i$, is a function of the Gaussian similarity between the node value and stimulus X.                   |
| Output Activation             | $O_j(X) = \sum_{k=1}^M w_{ji} \cdot a_i(X)$                                         | Activation of each Output unit $O_j$ is the weighted sum of the input activations and association weights.                             |
| Output Probability            | $P[Y_j|X] = \frac{O_j(X)}{\sum_{k=1}^M O_k(X)}$                                     | Each output node has associated response, $Y_j$. The probability of response $Y_j$ is determined by the ratio of output activations.   |
| Mean Output                   | $m(x) = \sum_{j=1}^L Y_j \cdot \frac{O_j(x)}{\sum_{k=1}^M O_k(X)}$                  | The response to stimulus x is the weighted average of the response probabilities.                                                      |
| **ALM Learning**              |                                                                                     |                                                                                                                                        |
| Feedback Activation           | $f_j(Z) = e^{-c(Z-Y_j)^2}$                                                          | After responding, feedback signal Z is presented, activating each output node via the Gaussian similarity to the ideal response.       |
| Update Weights                | $w_{ji}(t + 1) = w_{ji}(t) + \alpha \cdot (f_j(Z(t)) - O_j(X(t)) \cdot a_i(X(t))$   | Delta rule to update weights. Magnitude of weight changes controlled by learning rate parameter alpha.                                 |
| **EXAM**                      |                                                                                     |                                                                                                                                        |
| Extrapolation                 | $P[X_i|X] = \frac{a_i(X)}{\sum_{k=1}^M a_k(X)}$                                     | Novel test stimulus X activates input nodes associated with trained stimuli.                                                           |
|                               | $E[Y|X_i] = m(X_i) + \frac{m(X_{i+1})-m(X_{i-1})}{X_{i+1}-X_{i-1}} \cdot [X - X_i]$ | Slope value computed from nearest training instances and then added to the response associated with the nearest training instance,m(x) |

@@

Following the procedure used by Mcdaniel et al. (2009), we will assess the ability of both ALM and EXAM to account for the empirical data when fitting the models to 1) only the training data, and 2) both training and testing data. Models will be fit directly to the trial by trial data of each individual participants, both by minimizing the root-mean squared deviation (RMSE), and by maximizing log likelihood. Because ALM has been shown to do poorly at accounting for human patterns extrapolation (DeLosh et al., 1997), we will also fit the extended EXAM version of the model, which operates identically to ALM during training, but includes a linear extrapolation mechanism for generating novel responses during testing.

@@

hybridParamsC <-cbind(Model="Hybrid Test Only",pluck(hybrid_te_c, "Fit"), pluck(hybrid_te_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred))) |>
  rbind(cbind(Model="Hybrid Test & Train", pluck(hybrid_tetr_c,"Fit"), pluck(hybrid_tetr_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  rbind(cbind(Model="Hybrid Train Only", pluck(hybrid_tr_c, "Fit"), pluck(hybrid_tr_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

@@

@@

@@

```{r}
pluck(a_te_v, "train") |> pivot_longer(y:almResp, names_to="Resp", values_to = "vx") |> 
  mutate(dev=x-vx,abs_dev=abs(x-vx)) |>
  learn_curve_plot(tr, vx, Resp,facet_var=x, groupVec=Resp,nbins=8)

pluck(a_te_v, "train") |> pivot_longer(y:almResp, names_to="Resp", values_to = "vx") |> 
  mutate(dev=x-vx,abs_dev=abs(x-vx)) |>
  ungroup() %>%
  gather(key = "variable", value = "y_value", dev, abs_dev, vx) %>%
  group_by(variable) %>%
  group_map(~ learn_curve_plot(.x, x_var = tr, y_var = y_value, color_var = Resp, facet_var = x, groupVec = Resp, nbins = 8, y_label = .y$variable), .keep = TRUE)

list(a_tr_v, a_te_v,a_tetr_v) |> map( ~{pluck(.x, "train") |> pivot_longer(y:almResp, names_to="Resp", values_to = "vx") |> mutate(dev=x-vx,abs_dev=abs(x-vx)) |>
  ungroup() %>%
  gather(key = "variable", value = "y_value", dev, abs_dev, vx) %>%
  group_by(variable) %>%
  group_map(~ learn_curve_plot(.x, x_var = tr, y_var = y_value, color_var = Resp, facet_var = x, groupVec = Resp, nbins = 8, y_label = .y$variable), .keep = TRUE) })
```

# EXAM fit learning curves

```{r}
pluck(ex_te_v, "train") |> pivot_longer(y:almResp, names_to="Resp", values_to = "vx") |> 
  mutate(dev=x-vx,abs_dev=abs(x-vx)) |>
  learn_curve_plot(tr, vx, Resp,facet_var=x, groupVec=Resp,nbins=8)

pluck(ex_te_v, "train") |> pivot_longer(y:almResp, names_to="Resp", values_to = "vx") |> 
  mutate(dev=x-vx,abs_dev=abs(x-vx)) |>
  ungroup() %>%
  gather(key = "variable", value = "y_value", dev, abs_dev, vx) %>%
  group_by(variable) %>%
  group_map(~ learn_curve_plot(.x, x_var = tr, y_var = y_value, color_var = Resp, facet_var = x, groupVec = Resp, nbins = 8, y_label = .y$variable), .keep = TRUE)
```

```{r}
optimize_params_weighted_individual <- function(ds, c_values, lr_values, weight_exam_values, input.layer, output.layer) {
    all_results <- list()

    # Loop through each unique id
    for (individual in unique(ds$id)) {
        indiv_data <- ds[ds$id == individual, ]

        # Run the optimization function for the individual's data
        result <- optimize_params_weighted(indiv_data, c_values, lr_values, weight_exam_values, input.layer, output.layer)

        all_results[[as.character(individual)]] <- result
    }

    all_results
}

dss <- ds |> filter(id %in% c(1,2))

all_results_weighted_hybrid <- readRDS(here::here('data/model_cache/indv_hybrid_fits.rds'))
ma <- map(all_results_weighted_hybrid, "best_params") |> map("c")

data.frame(id=names(ma),c=as.numeric(ma))

ma = cbind(id=names(all_results_weighted_hybrid),map(all_results_weighted_hybrid, "best_params") |> map_dfr(magrittr::extract,c("c","lr","weight_exam")))

ds |> group_by(id,condit) |> distinct(id,condit) |> left_join(ma,by=join_by(id))

map(all_results_weighted_hybrid,"best_params") |> pluck("c")
all_results_weighted_hybrid[["1"]]$b

 map(~ map(.x$best_params, pluck, "c"))

 map_df(~ map_df(.x$train, pluck, "d"), .id = "density")
```

>>>>

# dsAvg |> filter(condit=="Constant",expMode2=="Test") %>% nrow(.)

# number of distinct bands in the Test mode

dsAvg |> filter(condit=="Varied",expMode2=="Test") %>% nrow(.)
dsAvg |> filter(condit=="Constant",expMode2=="Test") %>% nrow(.)
dsAvg |> filter(condit=="Varied",expMode2=="Test") %>% pull(x) |> n_distinct()
dsAvg |> filter(condit=="Constant",expMode2=="Test") %>% pull(x) |> n_distinct()

@@

| Step                          | Equation                                                                            | Description                                                                                                                            |
|------------------|------------------------|------------------------------|
| **ALM Activation & Response** |                                                                                     |                                                                                                                                        |
| Input Activation              | $a_i(X) = \frac{e^{-c(X-X_i)^2}}{\sum_{k=1}^M e^{-c(X-X_k)^2}}$                     | Activation of each input node $X_i$, is a function of the Gaussian similarity between the node value and stimulus X.                   |
| Output Activation             | $O_j(X) = \sum_{k=1}^M w_{ji} \cdot a_i(X)$                                         | Activation of each Output unit $O_j$ is the weighted sum of the input activations and association weights.                             |
| Output Probability            | $P[Y_j|X] = \frac{O_j(X)}{\sum_{k=1}^M O_k(X)}$                                     | Each output node has associated response, $Y_j$. The probability of response $Y_j$ is determined by the ratio of output activations.   |
| Mean Output                   | $m(x) = \sum_{j=1}^L Y_j \cdot \frac{O_j(x)}{\sum_{k=1}^M O_k(X)}$                  | The response to stimulus x is the weighted average of the response probabilities.                                                      |
| **ALM Learning**              |                                                                                     |                                                                                                                                        |
| Feedback Activation           | $f_j(Z) = e^{-c(Z-Y_j)^2}$                                                          | After responding, feedback signal Z is presented, activating each output node via the Gaussian similarity to the ideal response.       |
| Update Weights                | $w_{ji}(t + 1) = w_{ji}(t) + \alpha \cdot (f_j(Z(t)) - O_j(X(t)) \cdot a_i(X(t))$   | Delta rule to update weights. Magnitude of weight changes controlled by learning rate parameter alpha.                                 |
| **EXAM**                      |                                                                                     |                                                                                                                                        |
| Extrapolation                 | $P[X_i|X] = \frac{a_i(X)}{\sum_{k=1}^M a_k(X)}$                                     | Novel test stimulus X activates input nodes associated with trained stimuli.                                                           |
|                               | $E[Y|X_i] = m(X_i) + \frac{m(X_{i+1})-m(X_{i-1})}{X_{i+1}-X_{i-1}} \cdot [X - X_i]$ | Slope value computed from nearest training instances and then added to the response associated with the nearest training instance,m(x) |

@@

Following the procedure used by Mcdaniel et al. (2009), we will assess the ability of both ALM and EXAM to account for the empirical data when fitting the models to 1) only the training data, and 2) both training and testing data. Models will be fit directly to the trial by trial data of each individual participants, both by minimizing the root-mean squared deviation (RMSE). Because ALM has been shown to do poorly at accounting for human patterns extrapolation (DeLosh et al., 1997), we will also fit the extended EXAM version of the model, which operates identically to ALM during training, but includes a linear extrapolation mechanism for generating novel responses during testing.

We also fit a 3rd, hybrid model, which generates responses as a weighted sum of ALM and EXAM predictions. For the hybrid model, predictions are computed by first generating separate predictions from ALM and EXAM, and then combining them using the following equation: $\hat{y} = (1 - w) \cdot alm_pred + w \cdot exam_pred$, where $w$ is a third fit parameters that sets the relative contribution between the two models. For the grid search, the weight parameter is varied from 0 to 1, and the resulting RMSE is recorded.

Each model was fit to the data in 3 different ways. 1) To just the testing data, 2) Both the training and testing data, 3) Only the training data. In all cases, the model only updates its weights during the training phase, and the weights are frozen during the testing phase. In all cases, only the ALM model generates predictions during the training phase. For the testing phase, all 3 models are used to generate predictions.

The Value column of the table reflects the minimum RMSE achieved when fitting the model. The Test_RMSE column is the main performance indicator of interest, and represents the RMSE for just the testing data - thus for the Test Only fits, the Value and Test_RMSE columns are identical - given that the models were trained and tested on the exact some data sets.  For the Test & Train, and Train Only fits - the Test_RMSE can diverge from the fitting values.

@@

hybridParamsC <-cbind(Model="Hybrid Test Only",pluck(hybrid_te_c, "Fit"), pluck(hybrid_te_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred))) |>
  rbind(cbind(Model="Hybrid Test & Train", pluck(hybrid_tetr_c,"Fit"), pluck(hybrid_tetr_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  rbind(cbind(Model="Hybrid Train Only", pluck(hybrid_tr_c, "Fit"), pluck(hybrid_tr_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

calculate_AIC <- function(rmse, n, k) {
  return(n * log((rmse^2)) + 2 * k)
}

calculate_BIC <- function(rmse, n, k) {
  return(n * log((rmse^2)) + log(n) * k)
}

n <- dsAvg |> filter(expMode2=="Test") %>% pull(x) |> n_distinct()

# ALM AIC and BIC
almParamsV$AIC <- mapply(calculate_AIC, rmse = almParamsV$Test_RMSE, MoreArgs = list(n = n, k = 2))
almParamsV$BIC <- mapply(calculate_BIC, rmse = almParamsV$Test_RMSE, MoreArgs = list(n = n, k = 2))

# EXAM AIC and BIC
examParamsV$AIC <- mapply(calculate_AIC, rmse = examParamsV$Test_RMSE, MoreArgs = list(n = n, k = 2))
examParamsV$BIC <- mapply(calculate_BIC, rmse = examParamsV$Test_RMSE, MoreArgs = list(n = n, k = 2))

# Hybrid AIC and BIC
hybridParamsV$AIC <- mapply(calculate_AIC, rmse = hybridParamsV$Test_RMSE, MoreArgs = list(n = n, k = 3))
hybridParamsV$BIC <- mapply(calculate_BIC, rmse = hybridParamsV$Test_RMSE, MoreArgs = list(n = n, k = 3))

hybridParamsV

```

```{r}

calculate_metrics <- function(df, n, k) {
  df %>% 
    mutate(
      AIC = calculate_AIC(Test_RMSE, n, k),
      BIC = calculate_BIC(Test_RMSE, n, k)
    ) %>% 
    mutate(across(where(is.numeric), round, 3))
}

# Number of distinct testing positions
n <- dsAvg |> filter(expMode2=="Test") %>% pull(x) |> n_distinct()

# Create a list of all model data frames
model_list <- list(
  almParamsV = list(data = almParamsV, k = 2),
  examParamsV = list(data = examParamsV, k = 2),
  hybridParamsV = list(data = hybridParamsV, k = 3),
  almParamsC = list(data = almParamsC, k = 2),
  examParamsC = list(data = examParamsC, k = 2),
  hybridParamsC = list(data = hybridParamsC, k = 3)
)

# Apply the calculate_metrics function to each model data frame
model_results <- map(model_list, ~calculate_metrics(.x$data, n, .x$k))

# Extract the results back into separate data frames
list2env(model_results, envir = .GlobalEnv)

# The final tables are now in the global environment with the same names as before
# For example, you can view the updated ALM params for the Varied group with:
almParamsV
examParamsV
```

@@

```{r}

calculate_metrics <- function(df, n, k) {
  df %>%
    summarise(
      Test_RMSE = RMSE(y, pred),
      AIC = calculate_AIC(Test_RMSE, n, k),
      BIC = calculate_BIC(Test_RMSE, n, k)
    ) %>%
    mutate(across(where(is.numeric), round, 3))
}

n <- dsAvg |> filter(expMode2 == "Test") %>% pull(x) |> n_distinct()

# Bind the rows for each model and condition, then calculate AIC and BIC
almParams <- bind_rows(
  cbind(Model = "ALM Test Only", calculate_metrics(pluck(a_te_v, "test"), n, 2)),
  cbind(Model = "ALM Test & Train", calculate_metrics(pluck(a_tetr_v, "test"), n, 2)),
  cbind(Model = "ALM Train Only", calculate_metrics(pluck(a_tr_v, "test"), n, 2))
)

examParams <- bind_rows(
  cbind(Model = "EXAM Test Only", calculate_metrics(pluck(ex_te_v, "test"), n, 2)),
  cbind(Model = "EXAM Test & Train", calculate_metrics(pluck(ex_tetr_v, "test"), n, 2)),
  cbind(Model = "EXAM Train Only", calculate_metrics(pluck(ex_tr_v, "test"), n, 2))
)

hybridParams <- bind_rows(
  cbind(Model = "Hybrid Test Only", calculate_metrics(pluck(hybrid_te_v, "test"), n, 3)),
  cbind(Model = "Hybrid Test & Train", calculate_metrics(pluck(hybrid_tetr_v, "test"), n, 3)),
  cbind(Model = "Hybrid Train Only", calculate_metrics(pluck(hybrid_tr_v, "test"), n, 3))
)

# Example to show the result
head(hybridParams)
```

### Varied Parameter fits and RMSE

@@

### Constant Parameter fits and RMSE

@@
